Quality Definitions of Qualities

### 2021-09-30 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Completed installability and almost completed correctness

\usepackage{natbib}
\usepackage{booktabs}
\usepackage{amsmath,amsthm}
\usepackage{hyperref}
\hypersetup{
colorlinks=true,        % false: boxed links; true: colored links
linkcolor=red,          % color of internal links (change box color with linkbordercolor)
citecolor=blue,         % color of links to bibliography
filecolor=magenta,      % color of file links
urlcolor=cyan           % color of external links
}
\usepackage{bbding}

@@

%\oddsidemargin 0mm
%\evensidemargin 0mm
%\textwidth 160mm
%\textheight 200mm

@@

% Purpose

In the course of conducting and presenting our software engineering related
research we have often desired clear definitions for relevant software
qualities, like maintainability, usability, productivity, sustainability, etc.
Unfortunately, the definitions of qualities that we found in the literature were
not always satisfying.  The inadequacy of existing definitions became
particularly clear when we recently undertook an exercise measuring the
\lq\lq{}state of the practice\rq\rq{} in different software domains, including
the domains of seismology software \citep{SmithEtAl2018}, Geographic Information
Systems \citep{SmithEtAl2018_arXivGIS} and mesh generators
\citep{smith2016state}. When determining methods to quickly assess quality
across large samples of existing software, we found communication amongst
ourselves and with others challenging. We craved unambiguous and standardized
definitions to enable and inspire methods for measuring software qualities.  To
help ground our future research, and possibly help others in the same situation,
we have collected an exhaustive list of definitions.  From this list we have
produced our set of recommended software quality definitions.

In the current work, quality is defined as a measure of the excellence or worth
of an entity.  As is standard practice, we do not think of quality as a single
measure, but rather as a set of measures.  In the overview of managing software
quality provided by \citet[p 110]{VanVliet2000} the different measures are
called quality attributes, quality factors, and quality criteria.  In
\citet{ISO9126} they are called quality characteristics, subcharacteristics and
attributes.  To keep our presentation simple, and to not imply a classification
hierarchy that we don't intend, we will just use the term qualities to refer to
the different measures of quality.

Many taxonomies have been proposed for software qualities, as in
\citet{McCallEtAl1977, boehm1976quantitative, ISO9126}. Rather than aiming for a
complete coverage of the history of qualities, the wide variety of potential
qualities, or the relationships between them, we will instead take a practical
approach; we will focus on the qualities that we have found most relevant in our
research.  This means we will include the usual software engineering suspects,
like usability and maintainability.  Furthermore, since we also do research on
scientific computing software, we will also include scientifically relevant
qualities, like validatability and reproducibility relevant.

Given our respect for \citet[p.\ 15--33]{GhezziEtAl2003}, we used their list of
qualities as a starting point. Our list, shown in
Table~\ref{Tab_SoftwareQualities}, includes all of their qualities, except for
timeliness, which we excluded because the already included quality of
productivity implies the importance of timeliness.  To the list from
\citet{GhezziEtAl2003} we have added the practical quality of installability,
since our aforementioned studies on the state of the practice in different
software domains (for instance \citet{SmithEtAl2018_arXivGIS}) have shown us
that none of the other qualities matter if the software cannot be properly
installed.  Given our interest in scientific computing, we also added the
qualities of validatability and reproducibility. Since we were not able to find
much information in the literature on the quality of verifiability, we also
added the closely related quality of testability to our List.  The quality of
sustainability was added since this term has emerged as a popular goal for
software projects. The order of the qualities in
Table~\ref{Tab_SoftwareQualities} follow the order used in
\cite{GhezziEtAl2003}, which puts related qualities (like correctness and
reliability) together.  The order is roughly the order in which qualities are
considered.  We followed these same guidelines when inserting our additional
qualities into the list.

\begin{table}[h!]
  \begin{center}
  \begin{tabular}{lllll}
  \toprule
  & External & Internal & Product & Process \\  
  \midrule
  Installability & \Checkmark & & \Checkmark & \\
  Correctness & \Checkmark & \Checkmark & \Checkmark & \Checkmark \\
  Reliability & \Checkmark & & \Checkmark & \Checkmark \\
  Robustness & \Checkmark & & \Checkmark & \Checkmark \\     
  Performance & \Checkmark & & \Checkmark & \\
  Usability & \Checkmark & & \Checkmark & \Checkmark \\
  Verifiability & & \Checkmark & \Checkmark & \Checkmark \\
  Validatability & \Checkmark & & \Checkmark & \\     
  Testability & & \Checkmark & \Checkmark & \\
  Maintainability &  & \Checkmark & \Checkmark & \Checkmark\\
  Reusability & & \Checkmark & \Checkmark & \Checkmark\\
  Portability & & \Checkmark & \Checkmark & \\
  Understandability & & \Checkmark &  & \\     
  Interoperability & \Checkmark & & \Checkmark & \\
  Productivity &  &  &  & \Checkmark \\
  Visibility &  & &  & \Checkmark \\
  Reproducibility & \Checkmark & & \Checkmark & \\
  Sustainability & & \Checkmark & \Checkmark & \\      
  \bottomrule
\end{tabular}
\end{center}
\caption{Software Qualities}
\label{Tab_SoftwareQualities}
\end{table}

Table~\ref{Tab_SoftwareQualities} makes a distinction between external and
internal qualities.  External qualities are the qualities that are relevant to
the user of the software while internal qualities are those that are relevant to
developers \citep[p.\ 16]{GhezziEtAl2003}.  Although the users are not typically
interested in internal qualities, it is the internal qualities that influence
external quality.  External qualities are assessed using the produced software
and any user relevant documentation.  Internal qualities reference such things
as the code, documentation, test cases, and the software development process.
Correctness is shown as both an internal and external quality because it is
possible to judge the correctness of software and of the various internal
artifacts produced while developing software.  The external quality of
reliability is influenced by the internal quality of verifiability.  Since
validatability is related to getting the right requirements, it is shown as a
external quality.  Although one could judge the usability of internal
documentation, the quality of usability is considered as external, with the
quality of understandability introduced to cover the internal equivalent.
Filling in the table the way we want to use it.

Some qualities focus on the product, others on the process, and yet others on
both.  This is shown in Table~\ref{Tab_SoftwareQualities}.  Define process.
Define product. Process quality improves product quality.  The check marks are
for qualities that can be applied to a process, not the qualities that a process
improves.  Examples.  Performance isn't for process.  For process use
productivity.

% Scope

The topic of quality has been visited and revisited many times through the
course of software engineering.  Starting with McCall, Boehm.  Iso standards.
Taxonomies, relationships between qualities.  Rather than introduce distinctions
that aren't relevant, we'll just focus on qualities of interest. We are not
endeavoring to review all of the available literature, nor are we trying to
provide a complete framework for all qualities and the relationships between
them.  We are taking a more practical approach.  We have identified the
qualities that come up most in our research.  Not specifically emphasizing
scientific computing, but when a specific context is needed for a definition,
the scientific context will be used.  For instance, the ISO definition of
accuracy is defined in terms of precision, but in scientific computing accuracy
and precision are two different concepts.  Some are product and some are process
and some are both.  Some are internal and some are external.

\wss{List of definitions don't include textbooks that simply repeat
definitions from the literature.}

% Methodology
Methodology:

  \item Collect the list of qualities to define, based on the qualities that are
  listed in software engineering texts, such as Ghezzi Et al and Van Hliet.

  \item For each quality find as many definitions as possible

  \item Based on the existing definitions, propose a high quality definition of
  each quality.  The proposed definition may use one of the existing, or combine
  several of the options.  We may also make other modifications.  The goal is
  for the proposed definition to address our acceptance criteria.

software products, software artifacts and software development processes, and
ii) qualities that are considered important for good specifications.  The
specification could be a specification of requirements, design or a test plan.

The pattern in the presentation for each of the qualities is the same.  First we
summarize all of the definitions that we could find in the literature.  To keep
the clutter of quotation marks down, we have adopted the convention that each
definition is given verbatim from the cited source, but without showing
quotation marks.  In cases where the definition has to be rephrased, quotation
marks are used to show those portions that are taken verbatim for the original
source.  After the summary of the existing definitions, we propose the
definition that we would like to work with going forward.  This definition can
either be our preference from the existing definitions, or a new definition,
which is often found by combining existing definitions.  Following the proposed
definition is an explanation for the reasoning that led to this choice.

\wss{introduce the terminology of quality characteristics, as in ISO9126?}

\wss{tell the reader that the definitions will be sorted in the order of their
publication.}

In determining our recommended definition for each quality, we used the
following criteria:

@@

  \item[Unambiguous]

  \item[Consistency] The final list of definitions should be consistent in
  the terminology used.  For instance, we do not use synonyms for variety;
  we use the same word to mean the same concept throughout the final list of
  definitions.  This is why measurement is given by ``the effort required.''
  and not the various synonyms for this that are seen in the literature.
  \wss{We may change this phrase we use for measuring something.  I just
  wanted to put this in here as a placeholder.}

  \item[Generality]

  \item[Measurability]

  \item[Etc.]

@@

Not necessary to have every quality measurable.  Analogy to the definition of
true error in scientific computing.

Roadmap.
\subsection{Background}

Quality definitions history: McCall, Ghezzi et al, VanHleit, Boehm, ISO standards.

\section{Qualities of Software Products, Artifacts and Processes}
\label{SecQualities}

\wss{add a definition related to capability?  The word seems to come up in  many definitions.}

To assess the current state of software development, and to understand how
future changes impact software development, we need a clear definition of what
we mean by quality.  The concept of quality is decomposed into a set of separate
components that together make up ``quality''. Unfortunately, these are called
\emph{qualities}. These are associated to the software product, the software
artifacts (documentation, test cases, etc) and to the software development
process itself, and combinations thereof.

Our analysis is centred around a set of software qualities.  Quality is not
considered as a single measure, but a collection of different qualities, often
called ``ilities.''  These qualities highlight the desirable nonfunctional
properties for software artifacts, which include both documentation and
code. Some qualities, such as visibility and productivity, apply to the process
used for developing the software. The following list of qualities is based on
\cite{GhezziEtAl2003}. To the list from \cite{GhezziEtAl2003}, we have added
three qualities important for SC: installability, reproducibility and
sustainability.

@@

\subsection{Installability \oo{owner}}

\begin{defn}
  The capability of the software product to be installed in a specified
  environment \citep{ISO9126}. %2001
\end{defn}

@@

\noindent \textbf{Proposed Definition}

Combined definitions \ref{Installability_Selected2} and
\ref{Installability_Selected1} rephrased: Installability refers to the effort
required for the installation, uninstallation or reinstallation of a software or
product in a specified environment.

\wss{Remove the repetition of the word being defined in the definition?  Remove
  reinstallation?  This quality characteristic is a product quality.  It is an
  external quality.  Installability should be something that can be measured
  directly.}

\noindent \textbf{Reasoning}

 The definition captures a vital metric of measure for this quality - effort,
 and considers not just the installation but uninstallation and re-installation
 which are all related to installation activities.

 \wss{remove re-installation - it doesn't seem different from installation.
 Connection between installability and portability.  Use the terminology effort
 (without cost) to be consistent}

@@

 \subsection{Correctness \oo{owner}}

\begin{defn} 
  The extent to which a program satisfies its specifications and fulfills the
  user's mission objectives \citep{McCallEtAl1977}. 
\end{defn}

\begin{defn}
  The ability of software products to perform their exact tasks, as defined by
  their specification \citep{meyer1988object}.
\end{defn}

\begin{defn}  
  The degree to which a system or component is free from faults in its
  specification, design and implementation \citep{IEEEComputerDictionary1991}.
\end{defn}

\begin{defn} \label{Correctness_Selected}
  A program is functionally correct if it behaves according to its stated
  functional specifications \citep{GhezziEtAl2003}.  	
\end{defn}

\begin{defn}
  The degree to which a system is free from faults in its specification, design, and 
  implementation \citep{mcconnell2004code}.
\end{defn}

\noindent \textbf{Proposed Definition}

Definition~\ref{Correctness_Selected} rephrased: A program is correct if it
behaves according to its stated specifications.

\noindent \textbf{Reasoning}

To determine whether a program is correct, we must be able to refer to both its
specification, design and implementation to understand the intended purpose and
this definition captures it all.

@@

\subsection{\notdone{Reliability} \oo{owner}}

@@

\begin{defn} \label{reliabilitySelected2} Reliability is concerned with measuring
	the probability of occurence of faliure - that is, the probability of the
	observable effects of errors \citep{GhezziEtAl2003}.
\end{defn}

@@

\noindent \textbf{Proposed Definition}

Definition \ref{reliabilitySelected1} and \ref{reliabilitySelected2} rephrased:
Probability of failure-free operation of a computer program in a specified
environment for a specified time, i.e. the average time interval between two
failures also known as the mean time to failure (MTTF).

\wss{Needs to be completed.}\oo{Done}

\noindent \textbf{Reasoning}

Reliability is defined such that the measurement metrics are visible - the
specified environment and time.

@@

\subsection{Robustness \pmi{owner}}

@@

\noindent \textbf{Proposed Definition}

Definition \ref{RobustnessPartialDefn} and Definition
\ref{RobustnessDefnSelected} rephrased: Software possesses the characteristic of
robustness if it behaves ``reasonably'' in two situations: i) when it encounters
circumstances not anticipated in the requirements specification; and ii) when
the assumptions in its requirements specification are violated.

\noindent \textbf{Reasoning}

This definition indicates that robustness is related to the quality of
correctness (both within and outside of it).

\pmi{Still may want to refine what we mean by reasonable}

@@

\subsection{\notdone{Performance} \pmi{owner}}

@@

\noindent \textbf{Proposed Definition}

Combined definition \ref{PerformanceDefnSelected} and
\ref{PerformanceDefnSelected2}: The degree to which a system or component
accomplishes its designated functions within given constraints, such as speed
(database response times, for instance), throughput (transactions per second),
capacity (concurrent usage loads), and timing (hard real-time
demands). \wss{What is the difference between speed and timing?}

\noindent \textbf{Reasoning}

This definition offers a comprehensive list of constraints that are commonly
associated with software performance, such as speed, throughput, capacity, and
timing.

\pmi{I will check if Wiegers discusses the difference between speed and
timing.}

\pmi{Wiegers did not discuss this difference.}

\pmi{Speed - change over time; timing - related to specification: when
something should happen in relation to something else - state transitions}

\wss{If we need both speed and timing, we should define them.  We aren't
  including accuracy from the IEEE definition.  Is that on purpose?}

\pmi{We have decided not to include accuracy. SS will find difference b/w
speed and timing from a colleague.}

@@

\subsection{\notdone{Usability} \jc{owner}}

@@

\begin{defn}
  The capability of the software to be understood, learned, used and liked by
  the user, when used under specified conditions \cite{ISO9126}.
\end{defn}
\noindent \textbf{Proposed Definition}

\noindent \textbf{Reasoning}

>>>>

\usepackage{natbib}
\usepackage{booktabs}
\usepackage{amsmath,amsthm}
\usepackage{hyperref}
\hypersetup{
colorlinks=true,        % false: boxed links; true: colored links
linkcolor=blue,         % color of internal links (change box color with linkbordercolor)
citecolor=blue,         % color of links to bibliography
filecolor=magenta,      % color of file links
urlcolor=cyan           % color of external links
}
\usepackage{bbding}
\usepackage{tcolorbox}

% new tcolorbox environment
% #1: tcolorbox options
\newtcolorbox{mybox}[1][]
{
  colframe = yellow!50!red,
  colback  = white,
  width=\dimexpr\textwidth+12mm\relax,
  enlarge left by=-6mm,
  #1,
}

@@

@@

In the course of conducting and presenting our software engineering related
research we have often desired clear definitions for relevant software
qualities, like maintainability, usability, productivity, sustainability, etc.
Unfortunately, the definitions of qualities that we have found in the literature
are not always satisfying.  This inadequacy of existing definitions became
particularly clear when we recently undertook an exercise measuring the
\lq\lq{}state of the practice\rq\rq{} in different software domains, including
the domains of seismology software \citep{SmithEtAl2018}, Geographic Information
Systems \citep{SmithEtAl2018_arXivGIS} and mesh generators
\citep{smith2016state}. When determining methods to quickly assess quality
across large samples of existing software, we found communication amongst
ourselves and with others challenging. We craved unambiguous and standardized
definitions to enable and inspire methods for measuring software qualities.  To
help ground our future research, and possibly help others in the same situation,
we have collected an exhaustive list of definitions.  From this list we have
produced our set of recommended software quality definitions.

In the current work, software is defined as either programs, which run (execute)
on a computer, or libraries, which provide services to be used by programs. Both
programs and libraries are created using computer code, written in a programming
language.

Quality is defined as a measure of the excellence or worth of an entity.  As is
standard practice, we do not think of quality as a single measure, but rather as
a set of measures.  That is, quality is a collection of different qualities,
often called ``ilities.''  In the overview of managing software quality provided
by \citet[p 110]{VanVliet2000} the different measures are called quality
attributes, quality factors, and quality criteria.  In \citet{ISO9126} they are
called quality characteristics, sub-characteristics and attributes.  To keep our
presentation simple, and to not imply a classification hierarchy that we don't
intend, we will just use the term qualities to refer to the different measures
of quality.

The qualities of most interest to project managers and users are summarized in
the next subsection (\ref{Sec_IndirectQualities}).  Since these qualities cannot
generally be measured directly, they are characterized as indirect qualities.
The relevant indirect qualities are listed in this section, while being
characterized as external, internal, product or process qualities.  The next
subsection lists the relevant direct qualities that assist in achieving the
indirect qualities.  The final subsection presents the methodology that was
employed to summarize, analyze and define each quality.

\subsection{Indirect Qualities} \label{Sec_IndirectQualities}

Many taxonomies have been proposed for software qualities, as in
\citet{McCallEtAl1977, boehm1976quantitative, ISO9126}. Rather than aiming for
complete coverage of the history of qualities, the wide variety of potential
qualities, or the relationships between them, we will instead take a practical
approach; we will focus on the qualities that we have found most relevant in our
research.  This means we will include the usual software engineering suspects,
like usability and maintainability.  In addition, since we also do research on
scientific computing software, we will also include scientifically relevant
qualities, like validatability and reproducibility.

Given our respect for \citet[p.\ 15--33]{GhezziEtAl2003}, we used their list of
qualities as a starting point. Our list, shown in
Table~\ref{Tab_SoftwareQualities}, includes all of their qualities, except for
timeliness, which was excluded because the already included quality of
productivity overlaps with timeliness.  To the list from
\citet{GhezziEtAl2003} we have added the practical quality of installability,
since our aforementioned studies on the state of the practice in different
software domains (for instance \citet{SmithEtAl2018_arXivGIS}) have shown us
that none of the other qualities matter if the software cannot be properly
installed.  Given our interest in scientific computing, we also added the
qualities of validatability and reproducibility. Since we were not able to find
much information in the literature on the quality of verifiability, we also
added the closely related quality of testability.  The quality of sustainability
was added since this term has emerged as a popular goal for software projects.

The order of the qualities in Table~\ref{Tab_SoftwareQualities} follow the order
used in \cite{GhezziEtAl2003}, which puts related qualities (like correctness
and reliability) together.  Moreover, the order is roughly the order in which
qualities are considered in practice.  We followed these same guidelines when
inserting our additional qualities into the list.

\begin{table}[h!]
  \begin{center}
  \begin{tabular}{lllll}
  \toprule
  & External & Internal & Product & Process \\  
  \midrule
  Installability & \Checkmark & & \Checkmark & \\
  Correctness & \Checkmark & \Checkmark & \Checkmark & \Checkmark \\
  Reliability & \Checkmark & & \Checkmark & \Checkmark \\
  Robustness & \Checkmark & & \Checkmark & \Checkmark \\     
  Performance & \Checkmark & & \Checkmark & \\
  Usability & \Checkmark & & \Checkmark & \Checkmark \\
  Verifiability & & \Checkmark & \Checkmark & \Checkmark \\
  Validatability & \Checkmark & & \Checkmark & \Checkmark\\     
  Testability & & \Checkmark & \Checkmark & \\
  Maintainability &  & \Checkmark & \Checkmark & \Checkmark\\
  Reusability & & \Checkmark & \Checkmark & \Checkmark\\
  Portability & & \Checkmark & \Checkmark & \\
  Understandability & & \Checkmark & \Checkmark & \\     
  Interoperability & \Checkmark & & \Checkmark & \\
  Productivity &  &  &  & \Checkmark \\
  Visibility &  & &  & \Checkmark \\
  Reproducibility & \Checkmark & & \Checkmark & \\
  Sustainability & & \Checkmark & \Checkmark & \\      
  \bottomrule
\end{tabular}
\end{center}
\caption{Software Qualities}
\label{Tab_SoftwareQualities}
\end{table}

Table~\ref{Tab_SoftwareQualities} makes a distinction between external and
internal qualities.  External qualities are the qualities that are relevant to
the user of the software while internal qualities are those that are relevant to
developers \citep[p.\ 16]{GhezziEtAl2003}.  Although the users are not typically
interested in internal qualities, it is the internal qualities that influence
external quality.  External qualities are assessed using the produced software
and any user relevant documentation.  Internal qualities reference such things
as the code, documentation, test cases, and the software development process.
Correctness is shown as both an internal and external quality because it is
possible to judge the correctness of software and to judge the correctness of
the various internal artifacts produced while developing the software.  The
external quality of reliability is influenced by the internal quality of
verifiability. Since validatability is related to getting the right
requirements, it is shown as an external quality.  Although one could judge the
usability of internal documentation, the quality of usability is considered as
external, with the quality of understandability designated to cover the internal
equivalent of usability.

Some qualities focus on the product, others on the process, and yet others on
both, as shown in Table~\ref{Tab_SoftwareQualities}.  The term product refers to
both the external and internal products.  The external products are the
user-visible products, like the software itself and any user documentation.  The
internal products are the products (sometimes called work products or artifacts)
relevant to the developers, like requirements documentation, design
documentation, test cases, makefiles, etc.  The process refers to the sequence
of activities that produce the software.  The process generates both the
internal and external products.  The process is based on principles, methods and
techniques.  The process is supported by tools.  All product qualities (internal
and external) are effected by the process, so this information is not shown in
Table~\ref{Tab_SoftwareQualities}.  Instead, the checkmarks in
Table~\ref{Tab_SoftwareQualities} are for qualities that can be applied to a
process.  For instance, we do discuss the installability or correctness of a
process, but it does make sense to think about the portability of a process.
Rather than discuss the performance of a process, we introduce the process
specific quality of productivity.  Similarly, although we can discuss the
usability of a process, we instead introduce the term understandability to
exclusively apply to internal products (like the code).

The qualities shown in Table~\ref{Tab_SoftwareQualities} are qualities that can
only be measured indirectly \citep[p.\ 109]{VanVliet2000}.  For instance, a
direct measure of maintainability doesn't exist.  Since it cannot be measured
directly, we base our estimation of maintainability on direct measures that are
likely to influence maintainability, like consistency, traceability, etc.
Qualities that can be measured directly are discussed in the next section.

\subsection{Direct Qualities} \label{SecDirectQsIntro}

To achieve the qualities listed in Section~\ref{Sec_IndirectQualities},
conventional wisdom holds that the internal products should aim for the
qualities listed below.  All but the final quality listed (abstraction), are
adapted from the IEEE recommended practice for producing good software
requirements \citep{IEEE1998}. Abstraction is an important software development
principle for dealing with complexity \citep[p .~40]{GhezziEtAl2003}.
\citet{IEEE1998} also includes correctness and verifiability, but we have
excluded these from the list of direct qualities, since they belong in the list
of indirect qualities in the previous section.

\begin{itemize}
  \item Completeness
  \item Consistency
  \item Modifiability
  \item Traceability
  \item Unambiguity
  \item Abstract
\end{itemize}

All of the direct qualities listed above are internal qualities.  Furthermore,
they apply to both the (internal) product and the process.

\subsection{Methodology}

For each quality we collect as many distinct instances of the definition that we
can find.  Some of the definitions occur in multiple places (for instance the
definitions from \citet{McCallEtAl1977} appear in many software engineering
textbooks).  Since these instances do not add any new information, we do not
make an effort to identify how frequently the definitions reappear in different
sources.  Once all the definitions are collected, we determine the definition
that we would like to work with going forward.  This definition can either be
our preference from the existing definitions, or a new definition, which is
often found by combining existing definitions.

To ensure that our recommended definitions are quality definitions, they are
based on direct quality measures (\ref{SecDirectQsIntro}).  To the list of
direct qualities, we add the quality criteria of measurability.  How the direct
qualities are applied to develop definitions of qualities is described below.

@@

  \item[Completeness] To judge the completeness of our definitions, we verify
  that all of the relevant points from the collected definitions have been
  included.  We also verify that all facets of the definition are covered when a
  quality applies to multiple categories of internal, external, product and/or
  process.  Finally, we check that all of the qualities listed here have been
  considered.

  \item[Consistency] The final list of definitions should be consistent in the
  terminology used.  For instance, we do not use synonyms for variety; we use
  the same word to mean the same concept throughout the final list of
  definitions.  Specific decisions on terminology are as follows:
  \begin{itemize}
    \item To represent the idea of cost, effort, time, resources, and
    efficiency, we say \emph{effort}.
    \item To represent ``the degree to which'' or the ``the extent to which'',
    we will say \emph{the extent to which}.
    \item Fill in as go through definitions.
  \end{itemize}

  \item[Modifiability] The modifiability of the definitions is aided by the fact
  that the definitions are short.  Changes are not difficult to make.  The
  modifiability is further aided by the related qualities of traceability and
  abstraction.

  \item[Traceability] Each definition includes a section for the rationale.  The
  rationale explicitly links the recommended definition to the previously
  published definitions.  The rationale also explains why any aspects of the
  previous definitions were ignored.  The rationale also traces the selected
  definition to these qualities.

  \item[Unambiguity] With natural language it is difficult to remove all
  ambiguity, but every effort is made to keep the definitions simple and to use
  standard, consistent, terminology.

  \item[Abstract] Quality definitions should say what the quality achieves, but
  not how to achieve it.  We aim for abstract definitions that do not assume any
  specific methodology or tools.  For those qualities that fit into multiple
  categories of internal, external, product and/or process, we aim to have one
  definition that will apply to all.  When that is not possible, we will clarify
  how the definition is modified between categories.

  \item[Measurability] Although many qualities can only be indirectly measurable,
  the definition of the quality should imply a measure that is conceptually
  possible, even if the data needed to complete the measurement is often
  unavailable.  The scale type of the units of measure are as follows, in order
  of preference: ratio, interval, ordinal and nominal.  These scale types are
  defined in \citet[p.\ 107]{VanVliet2000}.  The ratio and interval scale types
  are preferred to the ordinal and nominal types because more they can convey
  more information.  Specific consequences of aiming for measurability are as follows:
  \begin{itemize}
    \item We will avoid the common phrase in definitions of ``the capability
    of'' because this implies a binary measure.  By this definition a product
    (or process) is either capable or not; it is a binary measure.  Instead of
    measuring capability, we will prefer to use the phrase ``the effort
    required''.
    \item etc.
  \end{itemize}

@@

Our aim is to capture the essence of the quality in an unambiguous way, even if
it cannot be measured. This is analogous to the definition of true error in
scientific computing.  The true error, which is the difference between the
calculated and the true solution, can rarely be calculated since the true
solution is rarely known. Even though it can usually only be estimated, the
concept of true error is still a concept that is relevant for analyzing and
understanding the behaviour of numerical algorithms.

After completing each definition we verify that the definition works for all of
the quality categories (internal, external, product and process) that apply.  If
not, the definition is modified.  The rationale section is written to explicitly
addresses each of these points.  Each definition is also checked to verify that
it is complete, consistent, traceable, unambiguous, abstract and measurable.
The rational section is written to explain how these qualities, for each quality
definition, are achieved.

@@

\section{Indirect Qualities of Software, Artifacts and Processes}
\label{SecQualities}

The presentation below for each of the qualities is the same.  First, we
summarize, in the order of their publication, all of the definitions that we
could find in the literature.  To keep the clutter of quotation marks down, we
have adopted the convention that each definition is given verbatim from the
cited source, but without showing quotation marks.  In cases where the
definition has to be rephrased, quotation marks are used to show those portions
that are taken verbatim for the original source.  Second, after the summary of
the existing definitions, we propose the definition that we would like to work
with going forward. Third, following the proposed definition is an explanation
for the reasoning that led to this choice.  The definitions apply to internal,
external, product and process following the designations in
Table~\ref{Tab_SoftwareQualities}. Where necessary, this is clarified in the
definition and its rationale.

\subsection{Installability (external, product)} %\oo{owner}

\begin{defn} \label{Installability_ISO9126}
  The capability of the software product to be installed in a specified
  environment \citep{ISO9126}. %2001
\end{defn}

@@

\begin{mybox}
\subsubsection*{Installability}

The effort required for the installation and/or uninstallation of software in a
specified environment.

\end{mybox}

\subsubsection*{Reasoning}

The recommended definition comes from combining Definitions
\ref{Installability_Selected2} and \ref{Installability_Selected1}.  The
different reasons for the proposed definition are as follows:

\begin{itemize}
  \item For consistency uses the term \textbf{effort} in place of efficiency.
  \item For measurability avoids the term ``the capability of'' from
  Definition~\ref{Installability_ISO9126}.
  \item The ``effectiveness'' from Definition~\ref{Installability_Selected2} is
  dropped because installation typically is judged as a binary result.  The
  installation is either completed, or not.
  \item From Definition~\ref{Installability_Selected1} the phrase ``given an
  installation is possible to begin with'' is removed because this notion is
  covered by the measure of effort.  The effort will be infinite if an
  installation is impossible.
\end{itemize}

@@

 \subsection{Correctness (external, internal, product, process)} %\oo{owner}

\begin{defn} \label{Correctness_McCallEtAl1977}
  The extent to which a program satisfies its specifications and fulfills the
  user's mission objectives \citep{McCallEtAl1977}. 
\end{defn}

\begin{defn} \label{Correctness_meyer1988object}
  The ability of software products to perform their exact tasks, as defined by
  their specification \citep{meyer1988object}.
\end{defn}

\begin{defn} \label{Correctness_IEEEComputerDictionary1991}
  The degree to which a system or component is free from faults in its
  specification, design and implementation \citep{IEEEComputerDictionary1991}.
\end{defn}

\begin{defn} \label{Correctness_GhezziEtAl2003}
  A program is functionally correct if it behaves according to its stated
  functional specifications \citep{GhezziEtAl2003}.  	
\end{defn}

\begin{defn} \label{Correctness_mcconnell2004code}
  The degree to which a system is free from faults in its specification, design,
  and implementation \citep{mcconnell2004code}.
\end{defn}

\begin{mybox}
\subsubsection*{Correctness}
An entity is correct if it matches its specification.  The entity can be either
software, an internal product (work product or artifact) or process.  The
specification can either be explicitly stated or implicit.
\end{mybox}

\subsubsection*{Reasoning}

The selection definition is a rephrasing of
Definition~\ref{Correctness_GhezziEtAl2003}. The different reasons for the
proposed definition are as follows:

\begin{itemize}
  \item Definition~\ref{Correctness_McCallEtAl1977} mentions ``mission
  objectives''.  We feel that the mission objectives are incorporated into the
  specification, so this element is not need to be mentioned separately in the
  recommended definition.
  \item Definition~\ref{Correctness_GhezziEtAl2003} refers to the functional
  specification, but we believe that software is incorrect if it does not
  follow the nonfunctional specification as well.  For instance, for safety
  critical software the performance requirements need to be satisfied, or the
  software will not be trustworthy.
  \item Definition~\ref{Correctness_IEEEComputerDictionary1991} and
  Definition~\ref{Correctness_mcconnell2004code} are almost the same, except
  Definition~\ref{Correctness_IEEEComputerDictionary1991} mentions ``system or
  component'' rather than just ``system''.  To keep our recommended definition
  simple, we do not introduce the distinction of components; we just focus on
  the entities of the software, internal products and processes.
  \item Several of the previous definitions imply measurement of different
  levels of correctness by phrases such as ``The extent to which''
  (Definition~\ref{Correctness_McCallEtAl1977}) and ``The degree to which''
  (Definitions~\ref{Correctness_IEEEComputerDictionary1991}
  and~\ref{Correctness_mcconnell2004code}). However, the proposed definition has
  a binary range of possible measurement - either correct or incorrect.  This
  doesn't follow our stated goal of having the quality of measurement cover the
  largest possible scale type.  We chose to recommend this simpler definition of
  correctness to satisfy the competing quality of unambiguity.
  Definition~\ref{Correctness_IEEEComputerDictionary1991} also implies a binary
  measure with the phrase ``the ability of''.  Making correctness a binary
  quality makes the definition unambiguous and it frees up room for the quality
  of reliability to cover the more nuanced aspects of the extent to which an
  entity is ``correct''.
  \item Internal Product,  Implied by
  Definition~\ref{Correctness_IEEEComputerDictionary1991}.  Changed to mention
  stated and unstated specification (explicit or implicit), since internal
  products do not have a specification.  May have a template.
  \item Process.
\end{itemize}

@@

\subsection{Reliability (external, product, process)} %\oo{owner} %NOTDONE

@@

\begin{defn} \label{reliabilitySelected2} Reliability is concerned with
	measuring the probability of occurrence of failure - that is, the probability
	of the observable effects of errors \citep{GhezziEtAl2003}.
\end{defn}

@@

\begin{mybox}
\subsubsection*{Reliability}
Probability of failure-free operation of a computer program in a specified
environment for a specified time, i.e. the average time interval between two
failures also known as the mean time to failure (MTTF).
\end{mybox}

\subsubsection*{Reasoning}

Definition \ref{reliabilitySelected1} and \ref{reliabilitySelected2} rephrased:.
Reliability is defined such that the measurement metrics are visible - the
specified environment and time.   The different reasons for the proposed
definition are as follows:

\begin{itemize}
  \item reason 1
  \item reason 2
  \item ...
\end{itemize}

@@

\subsection{Robustness (external, product, process)} %\pmi{owner}

@@

\begin{mybox}
\subsubsection*{Robustness}
Definition \ref{RobustnessPartialDefn} and Definition
\ref{RobustnessDefnSelected} rephrased: Software possesses the characteristic of
robustness if it behaves ``reasonably'' in two situations: i) when it encounters
circumstances not anticipated in the requirements specification; and ii) when
the assumptions in its requirements specification are violated.
\end{mybox}

\subsubsection*{Reasoning}

This definition indicates that robustness is related to the quality of
correctness (both within and outside of it). \pmi{Still may want to refine what
we mean by reasonable}  The different reasons for the proposed definition are as
follows:

\begin{itemize}
  \item reason 1
  \item reason 2
  \item ...
\end{itemize}

@@

\subsection{Performance (external, product)} %NOTDONE \pmi{owner}

@@

\begin{mybox}
\subsubsection*{Performance}
Combined definition \ref{PerformanceDefnSelected} and
\ref{PerformanceDefnSelected2}: The degree to which a system or component
accomplishes its designated functions within given constraints, such as speed
(database response times, for instance), throughput (transactions per second),
capacity (concurrent usage loads), and timing (hard real-time demands).
\wss{What is the difference between speed and timing?}
\end{mybox}

\subsubsection*{Reasoning}

This definition offers a comprehensive list of constraints that are commonly
associated with software performance, such as speed, throughput, capacity, and
timing. \pmi{I will check if Wiegers discusses the difference between speed and
timing.} \pmi{Wiegers did not discuss this difference.}  \pmi{Speed - change
over time; timing - related to specification: when something should happen in
relation to something else - state transitions}  \wss{If we need both speed and
timing, we should define them.  We aren't including accuracy from the IEEE
definition.  Is that on purpose?}  \pmi{We have decided not to include accuracy.
SS will find difference b/w speed and timing from a colleague.}   The
different reasons for the proposed definition are as follows:

\begin{itemize}
  \item reason 1
  \item reason 2
  \item ...
\end{itemize}

@@

\subsection{Usability (external, product, process) } %\notdone \jc{owner}

@@

\begin{defn}
  The capability of the software to be understood, learned, used and liked by
  the user, when used under specified conditions \cite{ISO9126}.
\end{defn}

\begin{mybox}
\subsubsection*{Usability}
The extent to which a product can be used by specified users to achieve
specified goals with effectiveness, efficiency, and satisfaction in a specified
context of use. \cite{ISO16982:2002} and \cite{ISO9241:11}
\end{mybox}

\subsubsection*{Reasoning}
